---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# inphr

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/tdaverse/inphr/graph/badge.svg)](https://app.codecov.io/gh/tdaverse/inphr)
<!-- badges: end -->

The goal of {inphr} is to provide a set of functions for performing null
hypothesis testing on samples of persistence diagrams using the theory of
permutations. Currently, only two-sample testing is implemented. Inputs can be
either samples of persistence diagrams themselves or vectorizations. In the
former case, they are embedded in a metric space using either the Bottleneck or
Wasserstein distance. In the former case, persistence data becomes functional
data and inference is performed using tools available in the
[{fdatest}](https://permaverse.github.io/fdatest/) package.

## Installation

You can install the development version of inphr from
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("tdaverse/inphr")
```

## Example

Let us start by loading the package:

```{r}
library(inphr)
```

The package contains three sets of persistence diagrams, which can be used for
testing. They are available in the package as `trefoils1`, `trefoils2`, and
`archspirals`. The first two sets contain persistence diagrams computed from
noisy samples of trefoil knots, while the third set contains persistence
diagrams computed from noisy samples of 2-armed Archimedean spirals. Each set
contains 24 persistence diagrams, each computed from a sample of 120 points
sampled from the respective shape, with Gaussian noise added (standard
deviation = 0.05). The persistence diagrams were computed using the
[`TDA::ripsDiag()`] function with a maximum scale of 6 and up to dimension 2.
You can use the `two_sample_test()` function to perform a two-sample test on
these persistence diagrams. For example, to test whether the first 5
persistence diagrams from the first set are significantly different from the
first 5 persistence diagrams from the second set, you can run:

```{r}
two_sample_test(trefoils1[1:5], trefoils2[1:5], B = 100L)
```

To test whether the first 5 persistence diagrams from the first set are
significantly different from the first 5 persistence diagrams from the third
set, you can run:

```{r}
two_sample_test(trefoils1[1:5], archspirals[1:5], B = 100L)
```

## Contributions

Contributions are welcome! Please feel free to open an issue or a pull request
if you have any suggestions or improvements. The package is still in its early
stages, so any feedback is appreciated.

### Code of Conduct

Please note that the {inphr} project is released with a [Contributor Code of
Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By
contributing to this project, you agree to abide by its terms.

### Acknowledgements

This project was funded by [an ISC grant from the R
Consortium](https://r-consortium.org/all-projects/2024-group-1.html#modular-interoperable-and-extensible-topological-data-analysis-in-r)
and done in coordination with Jason Cory Brunson and with guidance from Bertrand
Michel and Paul Rosen. It builds upon conversations with Mathieu Carrière and
Vincent Rouvreau who are among the authors of the
[GUDHI](https://gudhi.inria.fr) library. Package development also benefited
from the support of colleagues at the [Department of Mathematics Jean
Leray](https://www.math.sciences.univ-nantes.fr) and the use of equipment at
[Nantes University](https://english.univ-nantes.fr).
